---
title: "Research Question 2: Violent Crime by Reporting District in Los Angeles"
output: pdf_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Research Question 2

## Which areas in Los Angeles experience the highest rates of violent crime per district (Rpt Dist No), and how are these patterns distributed?

This analysis examines the geographic distribution of violent crime across Los Angeles reporting districts (`Rpt Dist No`).

Because population by reporting district is not available, the rate used in this analysis is defined as:

**violent crimes per 1,000 total reported crimes within each reporting district**

## Load libraries

```{r libraries}
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(knitr)
library(scales)
```

## Load datasets

```{r load-data}
# Full dataset of all reported crimes
all_crime_df <- read_csv("Fixed_Date_version.csv")

# Dataset already filtered to violent crimes
violent_df <- read_csv("Violent_Crimes.csv")
```

## Clean date columns

```{r clean-dates}
if (!inherits(all_crime_df$`DATE OCC`, "POSIXct") && !inherits(all_crime_df$`DATE OCC`, "Date")) {
  all_crime_df$`DATE OCC` <- parse_date_time(
    all_crime_df$`DATE OCC`,
    orders = c("Y b d I:M:S p", "Y-m-d H:M:S", "m/d/Y H:M", "m/d/Y I:M:S p")
  )
}

if (!inherits(violent_df$`DATE OCC`, "POSIXct") && !inherits(violent_df$`DATE OCC`, "Date")) {
  violent_df$`DATE OCC` <- parse_date_time(
    violent_df$`DATE OCC`,
    orders = c("Y b d I:M:S p", "Y-m-d H:M:S", "m/d/Y H:M", "m/d/Y I:M:S p")
  )
}
```

## Analysis from 2020–2023

```{r filter-years}
all_crime_df <- all_crime_df %>%
  mutate(year = year(`DATE OCC`)) %>%
  filter(year %in% 2020:2023)

violent_df <- violent_df %>%
  mutate(year = year(`DATE OCC`)) %>%
  filter(year %in% 2020:2023)
```

## Compute district-level totals and rates

- total number of crimes
- total number of violent crimes
- violent crime rate per 1,000 crimes
- percent of crimes that are violent

```{r district-summary}
district_total <- all_crime_df %>%
  group_by(`Rpt Dist No`, `AREA NAME`) %>%
  summarise(
    total_crime_count = n(),
    .groups = "drop"
  )

district_violent <- violent_df %>%
  group_by(`Rpt Dist No`, `AREA NAME`) %>%
  summarise(
    violent_count = n(),
    .groups = "drop"
  )

district_summary <- district_total %>%
  left_join(district_violent, by = c("Rpt Dist No", "AREA NAME")) %>%
  mutate(
    violent_count = ifelse(is.na(violent_count), 0, violent_count),
    violent_rate_per_1000 = (violent_count / total_crime_count) * 1000,
    violent_percent = (violent_count / total_crime_count) * 100
  ) %>%
  arrange(desc(violent_rate_per_1000))

kable(
  head(district_summary, 15),
  digits = 2,
  caption = "Top 15 Reporting Districts by Violent Crime Rate"
)
```

## Top districts

```{r top-districts}
top10_rate <- district_summary %>%
  slice_max(order_by = violent_rate_per_1000, n = 10)

top10_count <- district_summary %>%
  slice_max(order_by = violent_count, n = 10)

kable(top10_rate, digits = 2, caption = "Top 10 Districts by Violent Crime Rate")
kable(top10_count, digits = 2, caption = "Top 10 Districts by Violent Crime Count")
```

## Area-level summary

```{r area-summary}
area_summary <- district_summary %>%
  group_by(`AREA NAME`) %>%
  summarise(
    total_violent_crimes = sum(violent_count, na.rm = TRUE),
    total_crimes = sum(total_crime_count, na.rm = TRUE),
    avg_district_rate_per_1000 = mean(violent_rate_per_1000, na.rm = TRUE),
    median_district_rate_per_1000 = median(violent_rate_per_1000, na.rm = TRUE),
    district_count = n(),
    .groups = "drop"
  ) %>%
  mutate(
    area_violent_rate_per_1000 = (total_violent_crimes / total_crimes) * 1000
  ) %>%
  arrange(desc(area_violent_rate_per_1000))

kable(area_summary, digits = 2, caption = "Violent Crime Summary by LAPD Area")
```

## Visualization 1: Top 10 districts by violent crime rate

```{r plot-rate, fig.width=10, fig.height=6}
ggplot(
  top10_rate,
  aes(x = reorder(as.factor(`Rpt Dist No`), violent_rate_per_1000),
      y = violent_rate_per_1000,
      fill = `AREA NAME`)
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Reporting Districts by Violent Crime Rate",
    x = "Reporting District (Rpt Dist No)",
    y = "Violent crimes per 1,000 total crimes",
    fill = "Area"
  ) +
  theme_minimal()
```

## Visualization 2: Top 10 districts by violent crime count

```{r plot-count, fig.width=10, fig.height=6}
ggplot(
  top10_count,
  aes(x = reorder(as.factor(`Rpt Dist No`), violent_count),
      y = violent_count,
      fill = `AREA NAME`)
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Reporting Districts by Violent Crime Count",
    x = "Reporting District (Rpt Dist No)",
    y = "Violent crime count",
    fill = "Area"
  ) +
  theme_minimal()
```

## Visualization 3: Distribution of district violent crime rates

```{r plot-distribution, fig.width=9, fig.height=5}
ggplot(district_summary, aes(x = violent_rate_per_1000)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Violent Crime Rates Across Reporting Districts",
    x = "Violent crimes per 1,000 total crimes",
    y = "Number of districts"
  ) +
  theme_minimal()
```

## Visualization 4: Violent crime rate by area

```{r plot-area, fig.width=10, fig.height=6}
ggplot(
  area_summary,
  aes(x = reorder(`AREA NAME`, area_violent_rate_per_1000),
      y = area_violent_rate_per_1000)
) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  labs(
    title = "Violent Crime Rate by LAPD Area",
    x = "LAPD Area",
    y = "Violent crimes per 1,000 total crimes"
  ) +
  theme_minimal()
```

## Spatial distribution

This map is an approximate spatial summary using the median latitude and longitude of violent incidents within each district.

It is not a true district boundary map, but it helps show how high-violence districts are distributed geographically.

```{r map-data}
district_locations <- violent_df %>%
  filter(!is.na(LAT), !is.na(LON), LAT != 0, LON != 0) %>%
  group_by(`Rpt Dist No`) %>%
  summarise(
    median_lat = median(LAT, na.rm = TRUE),
    median_lon = median(LON, na.rm = TRUE),
    .groups = "drop"
  )

district_map <- district_summary %>%
  left_join(district_locations, by = "Rpt Dist No")
```

## Visualization 5: Approximate district map

```{r map-plot, fig.width=9, fig.height=7}
ggplot(
  district_map %>% filter(!is.na(median_lat), !is.na(median_lon)),
  aes(x = median_lon, y = median_lat,
      size = violent_count, color = violent_rate_per_1000)
) +
  geom_point(alpha = 0.75) +
  coord_quickmap(xlim = c(-118.7, -117.9), ylim = c(33.7, 34.4)) +
  labs(
    title = "Approximate Spatial Distribution of Violent Crime by Reporting District",
    x = "Longitude",
    y = "Latitude",
    size = "Violent count",
    color = "Rate per 1,000"
  ) +
  theme_minimal()
```

## Key results

```{r key-results}
highest_rate_district <- district_summary %>% slice_max(violent_rate_per_1000, n = 1)
highest_count_district <- district_summary %>% slice_max(violent_count, n = 1)
highest_area <- area_summary %>% slice_max(area_violent_rate_per_1000, n = 1)

cat("District with the highest violent crime rate:\n")
print(highest_rate_district)

cat("\nDistrict with the highest violent crime count:\n")
print(highest_count_district)

cat("\nArea with the highest violent crime rate:\n")
print(highest_area)
```

## Results summary

The analysis shows that violent crime is not evenly distributed across Los Angeles reporting districts. Instead, violent crime is concentrated in a smaller number of districts and areas.

Districts with the highest violent crime rate are not always the same districts with the highest violent crime count. This means that some districts have a larger share of violent crime relative to all reported crime, even if their total violent incident count is not the highest overall.

At the area level, some LAPD areas contain multiple districts with relatively high violent-crime rates, indicating geographic clustering.

The bar charts and histogram show that district-level violent crime varies substantially across Los Angeles, while the spatial point map shows that these high-rate districts are concentrated in particular locations rather than spread evenly across the city.

## Conclusion

Using violent crimes per 1,000 total crimes as the district-level rate measure, the results show that violent crime is spatially concentrated rather than evenly distributed. Some reporting districts consistently stand out as higher-risk districts, and some LAPD areas contain multiple high-rate districts.