---
title: "time analysis"
author: "Cailin Wilson"
date: "2026-05-04"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(readr)
library(lubridate)
```

Are certain types of crime more likely to occur at specific times of day or days of the week?

```{r, warning = FALSE}
df <- read_csv("Fixed_Date_version.csv")
```

```{r}
unique_codes <- unique(df[, c("Crm Cd", "Crm Cd Desc")])

unique_codes <- unique_codes[order(unique_codes$`Crm Cd`), ]

violent_codes <- c(
  110, 113, 121, 122, 210, 220, 230, 231, 235, 236,
  250, 251, 434, 435, 436, 437, 622, 623, 624, 625,
  626, 627, 647, 753, 755, 756, 760, 761, 763, 810,
  812, 815, 820, 821, 822, 860, 910, 920, 921, 922,
  926, 928, 930, 940, 943
)

violent_df <- df %>%
  mutate(
    crm_cd = as.numeric(`Crm Cd`)
  ) %>%
  filter(crm_cd %in% violent_codes)
```

```{r}
violent_df <- violent_df %>%
  mutate(
    
    # day of week
    date_only = as.Date(`DATE OCC`),

    weekday = wday(
      date_only,
      label = TRUE,
      abbr = FALSE
    ),

    # hour of day from TIME OCC
    time_chr = str_pad(
      as.character(`TIME OCC`),
      width = 4,
      pad = "0"
    ),

    hour = as.numeric(
      substr(time_chr, 1, 2)
    )
  )
```

```{r}
day_counts <- violent_df %>%
  count(weekday)

ggplot(day_counts,
       aes(x = weekday, y = n)) +
  geom_col() +
  labs(
    title = "Violent Crimes by Day of Week",
    x = "Day",
    y = "Count"
  )
```

```{r}
day_max <- day_counts %>%
  slice_max(n, n = 1)

day_min <- day_counts %>%
  slice_min(n, n = 1)

cat(
  "Highest:",
  as.character(day_max$weekday),
  "with",
  day_max$n,
  "violent crimes\n"
)

cat(
  "Lowest:",
  as.character(day_min$weekday),
  "with",
  day_min$n,
  "violent crimes\n"
)
```

```{r}
heat_data <- violent_df %>%
  count(weekday, hour)

ggplot(
  heat_data,
  aes(
    x = hour,
    y = weekday,
    fill = n
  )
) +
  geom_tile() +
  labs(
    title = "Violent Crime Patterns by Day and Hour",
    x = "Hour of Day",
    y = "Day of Week",
    fill = "Count"
  )
```

```{r}
peak_times <- violent_df %>%
  count(weekday, hour) %>%
  group_by(weekday) %>%
  slice_max(n, n = 1) %>%
  ungroup()

peak_times
```

```{r}
df_2020 <- df[df$`DATE OCC` < as.POSIXct("2021-01-01"),]

df_2021 <- df[df$`DATE OCC` < as.POSIXct("2022-01-01") & df$`DATE OCC` >= as.POSIXct("2021-01-01"),]

df_2022 <- df[df$`DATE OCC` < as.POSIXct("2023-01-01") & df$`DATE OCC` >= as.POSIXct("2022-01-01"),]

df_2023 <- df[df$`DATE OCC` < as.POSIXct("2024-01-01") & df$`DATE OCC` >= as.POSIXct("2023-01-01"),]

df_2024 <- df[df$`DATE OCC` < as.POSIXct("2025-01-01") & df$`DATE OCC` >= as.POSIXct("2024-01-01"),]
```

```{r}
violent_df2 <- df %>%
  mutate(
    crm_cd = as.numeric(`Crm Cd`)
  ) %>%
  filter(crm_cd %in% violent_codes) %>%
  mutate(
    
    date_only = as.Date(`DATE OCC`),

    year = year(date_only),

    month = month(
      date_only,
      label = TRUE,
      abbr = FALSE
    ),

    weekday = wday(
      date_only,
      label = TRUE,
      abbr = FALSE
    )
  )
```

```{r}
year_counts <- violent_df2 %>%
  count(year)

ggplot(year_counts,
       aes(x = factor(year), y = n)) +
  geom_col() +
  labs(
    title = "Violent Crimes by Year",
    x = "Year",
    y = "Count"
  )
cat("2024 appears to have a lower crime rate due to missing data")
```

```{r}
year_counts %>%
  arrange(desc(n))
```

```{r}
month_counts <- violent_df2 %>%
  count(month)

ggplot(month_counts,
       aes(x = month, y = n)) +
  geom_col() +
  labs(
    title = "Violent Crimes by Month",
    x = "Month",
    y = "Count"
  )
```

```{r}
peak_days_by_year <- violent_df2 %>%
  count(year, weekday) %>%
  group_by(year) %>%
  slice_max(n, n = 1) %>%
  ungroup()

peak_days_by_year
```

```{r}
lowest_days_by_year <- violent_df2 %>%
  count(year, weekday) %>%
  group_by(year) %>%
  slice_min(n, n = 1) %>%
  ungroup()

lowest_days_by_year
```

```{r}
heat_year_day <- violent_df2 %>%
  count(year, weekday)

ggplot(
  heat_year_day,
  aes(
    x = factor(year),
    y = weekday,
    fill = n
  )
) +
  geom_tile() +
  labs(
    title = "Violent Crime by Year and Weekday",
    x = "Year",
    y = "Day",
    fill = "Count"
  )
```

```{r}
weekday_year_table <- table(
  violent_df2$weekday,
  violent_df2$year
)

chisq.test(weekday_year_table)
```

A Pearson's chi-squared test was conducted to examine whether the distribution of violent crimes across days of the week differed by year in Los Angeles. Results indicated a statistically significant association between year and day of week, $\chi$ 2 (24)=60.97,p\<0.001, suggesting that while overall weekday crime patterns remained relatively consistent, the magnitude of violent crime across specific weekdays varied significantly across years.

```{r}
chi_result <- chisq.test(weekday_year_table)

chi2 <- as.numeric(chi_result$statistic)

n <- sum(weekday_year_table)

r <- nrow(weekday_year_table)

c <- ncol(weekday_year_table)

cramers_v <- sqrt(
  chi2 / (n * min(r - 1, c - 1))
)

cramers_v
```

Although weekday crime distributions differed significantly across years (p\<.001), the effect size was small, suggesting that temporal violent crime patterns in Los Angeles remained relatively stable over time.

```{r}
top_districts <- c(645, 1249, 1822, 119)

time_location_df <- violent_df %>%
 filter(`Rpt Dist No` %in% top_districts) %>%
 count(`Rpt Dist No`, hour)

```

```{r}
peak_points <- time_location_df %>%
  group_by(`Rpt Dist No`) %>%
  slice_max(n, n = 1)

ggplot(
  time_location_df,
  aes(
    x = hour,
    y = n,
    color = factor(`Rpt Dist No`)
  )
) +
  geom_line(size = 1) +
  geom_point(
    data = peak_points,
    size = 3
  ) +
  labs(
    title = "Hourly Violent Crime Patterns by Reporting District",
    x = "Hour of Day",
    y = "Violent Crime Count",
    color = "District"
  )
```

```{r}
district_extremes <- time_location_df %>%
  group_by(`Rpt Dist No`) %>%
  summarise(
    peak_hour = hour[which.max(n)],
    peak_count = max(n),

    low_hour = hour[which.min(n)],
    low_count = min(n)
  )

district_extremes
```

```{r}
heat_location <- violent_df %>%
 count(`AREA NAME`, weekday, hour)
ggplot(
 heat_location,
 aes(
   x = hour,
   y = weekday,
   fill = n
 )
) +
 geom_tile() +
 facet_wrap(~ `AREA NAME`)

```

```{r}
area_extremes <- heat_location %>%
  group_by(`AREA NAME`) %>%
  summarise(
    peak_day = weekday[which.max(n)],
    peak_hour = hour[which.max(n)],
    peak_count = max(n),

    low_day = weekday[which.min(n)],
    low_hour = hour[which.min(n)],
    low_count = min(n)
  )

area_extremes
```

```{r}
top_hotspots <- heat_location %>%
  arrange(desc(n)) %>%
  slice(1:5)

top_hotspots
```

```{r}
year_daily <- violent_df2 %>%
  count(year, date_only) %>%
  group_by(year) %>%
  summarise(
    avg_daily_crime = mean(n)
  )

ggplot(
  year_daily,
  aes(
    x = factor(year),
    y = avg_daily_crime
  )
) +
  geom_col() +
  labs(
    title = "Average Daily Violent Crimes by Year",
    x = "Year",
    y = "Average Daily Violent Crimes"
  )
```

```{r}
year_counts <- violent_df2 %>%
  count(year)

ggplot(
  year_counts,
  aes(
    x = year,
    y = n
  )
) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE)
```

```{r}
year_model <- lm(n ~ year, data = year_counts)

summary(year_model)
```

```{r}
top_areas <- c(
  "77th Street",
  "Central",
  "Hollywood",
  "Newton",
  "Southeast"
)

heat_top <- violent_df %>%
  filter(`AREA NAME` %in% top_areas) %>%
  count(`AREA NAME`, weekday, hour)

ggplot(
  heat_top,
  aes(
    x = hour,
    y = weekday,
    fill = n
  )
) +
  geom_tile() +
  facet_wrap(~ `AREA NAME`)
```

```{r}
heat_location <- violent_df %>%
  count(`AREA NAME`, weekday, hour)

top_areas <- violent_df %>%
  count(`AREA NAME`, sort = TRUE) %>%
  slice(1:4) %>%
  pull(`AREA NAME`)

heat_location_top <- violent_df %>%
  filter(`AREA NAME` %in% top_areas) %>%
  count(`AREA NAME`, weekday, hour)
```

```{r}
ggplot(
  heat_location_top,
  aes(
    x = hour,
    y = weekday,
    fill = n
  )
) +
  geom_tile() +
  facet_wrap(
    ~ `AREA NAME`,
    ncol = 2
  ) +
  labs(
    title = "Violent Crime Patterns in Top Hotspot Areas",
    x = "Hour of Day",
    y = "Day of Week"
  )
```

```{r}
area_extremes <- heat_location_top %>%
  group_by(`AREA NAME`) %>%
  summarise(
    peak_day = weekday[which.max(n)],
    peak_hour = hour[which.max(n)],
    peak_count = max(n),

    low_day = weekday[which.min(n)],
    low_hour = hour[which.min(n)],
    low_count = min(n)
  )

area_extremes
```
